---
title: "Problem Set 4"
author: "Zhengting (Johnathan) He"
date: "2022/3/12"
output:
  pdf_document: 
    toc_depth: 2
    latex_engine: lualatex
  html_document: default
  word_document: default
header-includes:
- \usepackage{amsmath,latexsym,amsfonts,amsthm,cleveref}
---


```{r "setup", include = FALSE}
require("knitr")
opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 4")
```


```{r}
setwd("D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 4")
require(tidyverse)
require(tidymodels)
require(poissonreg)
require(MASS)
require(gtsummary)

framData <- read_csv("./DATA/FraminghamPS4bin.csv")
```


**a. Explore the data using descriptive statistics for the 641 time bins (you do not have the individual data in this binned data set):**


```{r}
# summary statistics of all variables
skimr::skim(framData)
```


```{r}
# summary statistics of death rate
summary(framData$Rate)
```


```{r, results = "asis"}
# explore pairwise relationships between independent variable
# and binary mortality rate
framData %>% 
    mutate(Rate_bi = ifelse(Rate > 0.0000176, "high", "low")) %>% 
    mutate(Rate_bi = factor(Rate_bi)) %>% 
    dplyr::select(gender:agecat, Rate_bi) %>%
    tbl_summary(by = "Rate_bi") %>% 
    add_overall() %>% 
    add_p()
```


```{r}
# explore pairwise relationships between independent variable
# and binary mortality rate
framData %>% 
    mutate(Rate_bi = ifelse(Rate > 0.0000176, "high", "low")) %>% 
    mutate(Rate_bi = factor(Rate_bi)) %>% 
    dplyr::select(gender:agecat, Rate_bi) %>%
    pivot_longer(cols = - Rate_bi) %>%
    ggplot(aes(y = value, fill = Rate_bi)) +
    geom_bar(position = "fill") +
    facet_wrap(vars(name), scales = "free", ncol = 2) +
    labs(x = NULL, y = NULL, fill = NULL) +
    theme_minimal()
```


As shown from the output above, all independent variables except for `cursmoke` are significantly different between the two binary mortality rate groups.


```{r}
framData %>% 
    dplyr::select(gender:agecat) %>%
    mutate(across(everything(),
                  ~ as.numeric(.x))) %>% 
    corrr::correlate() %>% 
    corrr::rplot(colours = c("indianred2", "black", "skyblue1")) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
```


As shown the output above, `bpmeds` and `diabetes` are colinear.


```{r}
# Death rate across time bin
framData %>% 
    ggplot(aes(x = tbin, y = Rate, color = tbin)) +
    geom_jitter() +
    theme_minimal()
```


As shown from the output above, death rate may be different across time bin.


```{r, results = "asis"}
# explore distributions of baseline covariates across time bins
framData %>% 
    dplyr::select(gender:agecat, tbin) %>%
    tbl_summary(by = "tbin") %>% 
    add_overall() %>% 
    add_p()
```


**b. Explore several Poisson regression models using these grouped survival data and select between models:**


```{r}
# specify Poisson model
poisson_spec <- poisson_reg() %>% 
    set_engine("glm") %>% 
    set_mode("regression")
```


**Model 1:**


$$
log(D) \sim log(Y) + gender + diabetes + bmicat + agecat
$$


```{r}
model1 <- poisson_spec %>% 
    fit(D ~ gender + diabetes + bmicat + agecat + offset(log(Y)),
        data = framData)
```


```{r}
model1 %>% tidy() %>% kable()
```


```{r}
model1 %>% glance() %>% kable()
```


**Model 2:**


$$
log(D) \sim log(Y) + gender + bpmeds + bmicat + agecat
$$


```{r}
model2 <- poisson_spec %>% 
    fit(D ~ gender + bpmeds + bmicat + agecat + offset(log(Y)),
        data = framData)
```


```{r}
model2 %>% tidy() %>% kable()
```


```{r}
model2 %>% glance() %>% kable()
```


**Model 3:**


$$
log(D) \sim log(Y) + gender + diabetes + bpmeds + bmicat + agecat
$$


```{r}
model3 <- poisson_spec %>% 
    fit(D ~ gender + diabetes + bpmeds + bmicat + agecat + offset(log(Y)),
        data = framData)
```


```{r}
model3 %>% tidy() %>% kable()
```


```{r}
model3 %>% glance() %>% kable()
```


Based on `AIC` statistics, even though `diabetes` and `bpmeds` may be correlated, including both independent variables does improve model fit.


**Model 4:**


$$
log(D) \sim log(Y) + gender + diabetes + bpmeds + cursmoke + bmicat + agecat
$$


```{r}
model4 <- poisson_spec %>% 
    fit(D ~ gender + diabetes + bpmeds + cursmoke + bmicat + agecat +
            offset(log(Y)),
        data = framData)
```


```{r}
model4 %>% tidy() %>% kable()
```


```{r}
model4 %>% glance() %>% kable()
```


Based on `AIC` statistics, including `cursmoke` further improves model fit.


**Model 5:**


$$
log(D) \sim log(Y) + tbin + gender + diabetes + bpmeds + cursmoke + bmicat + agecat
$$


```{r}
model5 <- poisson_spec %>% 
    fit(D ~ tbin + gender + diabetes + bpmeds + cursmoke + bmicat +
            agecat + offset(log(Y)),
        data = framData)
```


```{r}
model5 %>% tidy() %>% kable()
```


```{r}
model5 %>% glance() %>% kable()
```


Based on `AIC` statistics, including `tbin` as continuous variable further improves model fit.


**Model 6:**


$$
log(D) \sim log(Y) + [tbin1 + tbin2 + tbin3 + tbin4 + tbin5] + gender + diabetes + bpmeds + cursmoke + bmicat + agecat
$$


```{r}
model6 <- poisson_spec %>% 
    fit(D ~ factor(tbin) + gender + diabetes + bpmeds + cursmoke +
            bmicat + agecat + offset(log(Y)),
        data = framData)
```


```{r}
model6 %>% tidy() %>% kable()
```


```{r}
model6 %>% glance() %>% kable()
```


As shown from the output above, including `tbin` as categorical variable improves the likelihood, however the `AIC` statistics is higher, since the `AIC` statistics panalize model with larger number of independent variables.


**Summary of all models tried:**


```{r, results = "asis"}
tbl_merge(list(model1, model2, model3, model4, model5, model6) %>% 
              map(~ .x %>% tbl_regression(exponentiate = TRUE)),
          tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**",
                          "**Model 4**", "**Model 5**", "**Model 6**")) %>% 
    bold_labels()
```


```{r}
list(model1, model2, model3, model4, model5, model6) %>% 
    map_dfr(~ glance(.x)) %>% 
    kable()
```


Based on `AIC` statistics, we may choose **model 5**.


**c. Check the assumptions of your Poisson models; use other models as appropriate:**


```{r}
model5 <- model5 %>% extract_fit_engine()
```


```{r}
# Pearson chi-square goodness-of-fit test (like poisgof in Stata)
X2 = sum(residuals(model5, type = "pearson")^2); X2
df = model5$df.residual; df
pval = 1-pchisq(X2, df); pval
```


```{r}
model5_nb <- glm.nb(D ~ tbin + gender + diabetes + bpmeds + cursmoke +
                        bmicat + agecat + offset(log(Y)),
                    data = framData)
summary(model5_nb)
```


```{r}
# likelihood ratio test compare Poisson with NB
lmtest::lrtest(model5, model5_nb) %>% tidy() %>% kable()
```


As shown in the goodness-of-fit test above, the Poisson model fits the data well. As shown from the likelihood ratio test above, we fail to reject the null hypothesis of $H_0: \alpha = 0$, indicating that the assumption of mean = variance of Poisson model holds.


**d. Save your R script file that documents and archives the steps of your statistical analysis. This file will make your analysis “reproducible.”**


**e. Summarize your findings in a brief report (less than two pages with at most one table and one figure) as if for a biomedical/public health journal.**


**A suggested format is:**

+ **Introduction – a few sentences about the research question(s)**

+ **Data description – simple tabulations describing the distributions of characteristics by time bins**

+ **Results from multiple models that address question(s) (e.g., bivariate and multivariable)**

+ **Graphical display that presents evidence in the data relevant to your scientific question.**


**Introduction**

In this problem set, we evaluate the relationship between mortality and baseline covariates using the Framingham Heart Study, which is a long term prospective study of the etiology of cardiovascular disease among a population of free living subjects in the community of Framingham, Massachusetts. To analyze the mortality rate, the data was grouped into 5-years intervals.


**Data description**


As shown from the table below, there are differences in the distribution of baseline gender, current smoking status, diabetic status, use of anti-hypertensive medications, body mass index category, and age category, across time bins; however, the differences are not statistically significant based on Chi-squared test.


```{r, echo = FALSE, results = "asis"}
framData %>% 
    dplyr::select(gender:agecat, tbin) %>%
    tbl_summary(by = "tbin") %>% 
    add_overall() %>% 
    add_p()
```


**Results**


As shown from the table below, we tried several models to explore the relationship between mortality and baseline covariates using Poisson regression, and select models based on the `AIC` statistics. In the final model selected (Model 5), the incidence rate ratio comparing female to male is 0.57 (95% CI: 0.51, 0.63); the incidence rate ratio comparing diabetes to non diabetes is 2.54 (95% CI: 2.03, 3.13); the incidence rate ratio associated with one unit increase in BMI is 1.15 (95% CI: 1.07, 1.24); the incidence rate ratio associated with one unit increase in age category is 2.24 (95% CI: 2.11, 2.38); the incidence rate ratio comparing current anti-hypertensive medications to not current users is 2.08 (95% CI: 1.67, 2.56); the incidence rate ratio comparing current smokers to not current smokers is 1.47 (95% CI: 1.32, 1.64). All incidence rate ratios are statistically significant under $\alpha = 0.05$.


```{r, echo = FALSE}
tbl_merge(list(model1, model2, model3, model4, model5, model6) %>% 
              map(~ .x %>% tbl_regression(exponentiate = TRUE)),
          tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**",
                          "**Model 4**", "**Model 5**", "**Model 6**")) %>% 
    bold_labels()
```


Based on these evidence, we suggest that male, diabetes, higher BMI, older age, anti-hypertensive medications usage, and smoking may associated with a higher rate of mortality.


**Graphic display**


```{r}
(model5 %>% tbl_regression())$table_body %>%
    select(variable, estimate, conf.low, conf.high) %>%
    mutate(estimate = exp(estimate),
           conf.low = exp(conf.low),
           conf.high = exp(conf.high)) %>% 
    slice(2:7) %>% 
    ggplot(aes(x = variable, y = estimate)) +
    geom_point() +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high, width = 0.05)) +
    geom_hline(aes(yintercept = 1)) +
    xlab("Independent variable") +
    ylab("Adjusted incidence rate ratio") +
    theme_minimal()
```

