---
title: "Problem Set 1"
author: "Zhengting (Johnathan) He"
date: "2022/1/28"
output: html_document
---


```{r "setup", include = FALSE}
require("knitr")
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 1")
```


```{r, message = FALSE}
# set up
require(tidyverse)
require(gtsummary)

setwd("D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 1")

data <- read_csv("./data/nepal621.csv")
```


**1. Suppose you are interested in choosing an appropriate sample size for estimating the 
16- month mortality rate for children younger than 3 years of age in a developing country in which vitamin A supplementation is not currently available.**


**a. Use the available information from the Nepal data set to choose a sample size so that you estimate this rate to within +/- 0.5%.**


---

```{r, results = "asis"}
data %>% 
    filter(age == "<1" | age == "1-2") %>% 
    filter(trt == "Placebo") %>% 
    tbl_summary(include = "status",
                digits = list(status ~ 2))
```

---

According to the summary statistics above, estimated $p = \hat{p} = 0.0294$, estimated $q = \hat{q} = 0.9706$.

$n = \frac{(z_{\alpha/2})^2 * pq}{d^2} = \frac{1.96^2*0.0294*0.9706}{0.005^2} = 4384.90 \approx 4385$.

---


**b. Now, suppose *no information* is available from this Nepal study. Determine what sample
size would be required for each of a range of plausible values of the mortality rate.
Summarize your sample size findings in a *table*.**


---

```{r}
seq(0.05, 0.95, 0.05) %>% 
    map_dfr(~ tibble(`Mortality rate` = .x,
                     `Sample size (N)` = ceiling(qnorm(0.975)^2 * (.x) * (1 - .x) / 0.005^2))) %>% 
    kable()
```

---

**2. Now suppose you have a chance to investigate the effect of vitamin A supplementation
on the mortality of children under 3 years of age. The `power.prop.test()` command in
R can be used with the results of the Nepal trial to choose the size of the vitamin A and
control groups (assuming equal sample sizes for both groups) for the new study. Confirm
from the data set that the 16-month mortality in the placebo group is 0.0294 and the 16-
month mortality in the Vitamin A group is 0.0245 for the Nepal study. The estimated
relative risk of death in the placebo group as compared to the Vitamin A group is
0.029/0.0245 = 1.2. Assuming a significance level of 0.05 and power of 80%, the sample
size needed in the new study to detect a relative risk of 1.2 is *17,144 children per group*
according to the results on the next page. A total sample size of 34,288 children would be
required.**


```{r}
power.prop.test(n = NULL, p1 = 0.0294, p2 = 0.0245, sig.level = 0.05, power = 0.8,
                alternative = "two.sided")
```


**3. Verify R’s calculations for part 2) by hand using the method learned in class. Expect your
answer to be close in value to, but not exactly the same as, that provided by `R`, due to
rounding in hand calculations. (`Stata` uses a continuity correction that `R` doesn’t, so the
value calculated from `Stata` will also be different than the one you calculated by hand and in
R.)**


---

According to the information provided, $p_1 = 0.0294$, $p_2 = 0.0245$, $\alpha = 0.05$, $\beta = 0.2$,
$\Delta = 0.0245 * (1.2 - 1) = 0.0049$, $\bar{p} = \frac{p_1 + p_2}{2} = 0.02695$.

$n_1 = n_2 = \frac{(z_{\alpha/2}*\sqrt{2*\bar{p}*\bar{q}} + z_{\beta}*\sqrt{p_1*q_1 + p_2*q_2})^2}{\Delta^2} = \frac{(1.96*\sqrt{2*(0.02695)*(1-0.02695)} + 0.84*\sqrt{(0.0294)*(1-0.0294) + (0.0245)*(1-0.0245)})^2}{(0.0049)^2} = 17124.5 \approx 17125$.

---

To avoid differences caused by rounding issues, I used `qnorm()` function to conduct the calculation:

```{r}
(qnorm(0.025) * sqrt(2 * (0.02695) * (1 - 0.02695)) + qnorm(0.2) * sqrt((0.0294) *
(1 - 0.0294) + (0.0245) * (1 - 0.0245)))^2/(0.0049^2)
```

---


**4. Construct a *table* that displays the total sample sizes required under various
assumptions about the mortality rate in the control group and the relative risk of interest.
Assume a significance level of 0.05 and 80% power. Comment on what you observe.**


**Vary the assumptions by:**

+ **a. Assuming that the control group mortality rate (risk) is:**

  - **1. the same as that observed in Nepal placebo group of children < 3 years of age**
  
  - **2. or .5% lower**
  
  - **3. or .5% higher**
  
+ **b. Assuming that the relative risk of death for children in the control group as compared
to children receiving vitamin A is hypothesized to be:**

  - **1. 1.2 (the same as the relative risk that was estimated for Nepali children in this age group**
  
  - **2. or 1.5**
  
  - **3. or 1.75.**
  

---

```{r}
map2_dfr(c(0.0294, 0.0294, 0.0294, 0.0244, 0.0244, 0.0244, 0.0344, 0.0344, 0.0344),
         c(1.2, 1.5, 1.75, 1.2, 1.5, 1.75, 1.2, 1.5, 1.75),
         ~ tibble(
             `Mortality Rate` = .x,
             `Relative Risk` = .y,
             `Sample Size` = ceiling(power.prop.test(
                  n = NULL,
                  p1 = .x,
                  p2 = .x/.y,
                  sig.level = 0.05,
                  power = 0.8,
                  alternative = "two.sided"
             )$n*2)
         )) %>% 
    pivot_wider(names_from = `Mortality Rate`,
                values_from = `Sample Size`) %>% 
    rename("Control Mortality Rate Same" = "0.0294",
           "Control Mortality Rate 0.5% Lower" = "0.0244",
           "Control Mortality Rate 0.5% Higher" = "0.0344") %>% 
    kable()
```

*Note: the sample size in the table above are the total sample size (i.e., sample size in both Vitamin A and placebo groups).*

---

With else factors ($\alpha$, $\beta$, $\text{relative risk}$) holding constant, sample size needed increases as mortality rate in placebo group decreases (the absolute risk difference between Vitamin A group and placebo group decreases).

With else factors ($\alpha$, $\beta$, $\text{mortality rate in control group}$) holding constant, sample size needed decreases as relative risk of placebo group compared to treatment group increases.

---


**5. Construct another *table* that displays the total sample sizes required under the same 
varying assumptions of the mortality rate in the control group and the relative risk of interest. 
This time, assume a significance level of 0.05 and 90% power. Comment on what you 
observe.**


---

```{r}
map2_dfr(c(0.0294, 0.0294, 0.0294, 0.0244, 0.0244, 0.0244, 0.0344, 0.0344, 0.0344),
         c(1.2, 1.5, 1.75, 1.2, 1.5, 1.75, 1.2, 1.5, 1.75),
         ~ tibble(
             `Mortality Rate` = .x,
             `Relative Risk` = .y,
             `Sample Size` = ceiling(power.prop.test(
                  n = NULL,
                  p1 = .x,
                  p2 = .x/.y,
                  sig.level = 0.05,
                  power = 0.9,
                  alternative = "two.sided"
             )$n*2)
         )) %>% 
    pivot_wider(names_from = `Mortality Rate`,
                values_from = `Sample Size`) %>% 
    rename("Control Mortality Rate Same" = "0.0294",
           "Control Mortality Rate 0.5% Lower" = "0.0244",
           "Control Mortality Rate 0.5% Higher" = "0.0344") %>% 
    kable()
```

*Note: the sample size in the table above are the total sample size (i.e., sample size in both Vitamin A and placebo groups).*

---

With else factors ($\alpha$, $\beta$, $\text{relative risk}$) holding constant, sample size needed increases as mortality rate in placebo group decreases (the absolute risk difference between Vitamin A group and placebo group decreases).

With else factors ($\alpha$, $\beta$, $\text{mortality rate in control group}$) holding constant, sample size needed decreases as relative risk of placebo group compared to treatment group increases.

Compared to results in 4., with else factors ($\alpha$, $\text{relative risk}$, $\text{mortality rate in control group}$) holding constant, sample size needed increases as power increases.

---

**6. Select a design based upon your findings from parts 4 and 5 above. Write a brief paragraph 
that presents and justifies your choice. Be numerate.**

---

We consider conducting a superiority trial to compare Vitamin A supplement to placebo on reducing the risk of mortality in children younger than 3 years of age in a developing country. Based on a previous study in similar settings, we hypothesize the relative risk of mortality in the control group compared to Vitamin A group as 1.2, and the mortality rate in placebo group is around 2.94%. Thus, under a significant level of 0.05 and a statistical power of 0.8, at least 34288 total participants (17144 in each group) are needed to observe a significant difference. When further considering a drop-off rate of 10%, 38098 total participants (19049 in each group) are needed.

---

