---
title: "Problem Set 1"
author: "Zhengting (Johnathan) He"
date: "2022/1/28"
output: html_document
---


```{r "setup", include = FALSE}
require("knitr")
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 1")
```


```{r, message = FALSE}
# set up
require(tidyverse)
require(gtsummary)

setwd("D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 1")

data <- read_csv("./data/nepal621.csv")
```


**1. Suppose you are interested in choosing an appropriate sample size for estimating the 
16- month mortality rate for children younger than 3 years of age in a developing country in which vitamin A supplementation is not currently available.**


**a. Use the available information from the Nepal data set to choose a sample size so that you estimate this rate to within +/- 0.5%.**


---

```{r, results = "asis"}
data %>% 
    filter(age == "<1" | age == "1-2") %>% 
    filter(trt == "Placebo") %>% 
    tbl_summary(include = "status",
                digits = list(status ~ 2))
```

---

According to the summary statistics above, estimated $p = \hat{p} = 0.0294$, estimated $q = \hat{q} = 0.9706$.

$n = \frac{(z_{\alpha/2})^2 * pq}{d^2} = \frac{1.96^2*0.0294*0.9706}{0.005^2} = 4384.90 \approx 4385$.

---


**b. Now, suppose *no information* is available from this Nepal study. Determine what sample
size would be required for each of a range of plausible values of the mortality rate.
Summarize your sample size findings in a *table*.**


---

```{r}
seq(0.05, 0.95, 0.05) %>% 
    map_dfr(~ tibble(`Mortality rate` = .x,
                     `Sample size (N)` = ceiling(qnorm(0.975)^2) * (.x) * (1 - .x) / 0.005^2)) %>% 
    kable()
```

---

**2. Now suppose you have a chance to investigate the effect of vitamin A supplementation
on the mortality of children under 3 years of age. The `power.prop.test()` command in
R can be used with the results of the Nepal trial to choose the size of the vitamin A and
control groups (assuming equal sample sizes for both groups) for the new study. Confirm
from the data set that the 16-month mortality in the placebo group is 0.0294 and the 16-
month mortality in the Vitamin A group is 0.0245 for the Nepal study. The estimated
relative risk of death in the placebo group as compared to the Vitamin A group is
0.029/0.0245 = 1.2. Assuming a significance level of 0.05 and power of 80%, the sample
size needed in the new study to detect a relative risk of 1.2 is *17,144 children per group*
according to the results on the next page. A total sample size of 34,288 children would be
required.**


```{r}
power.prop.test(n = NULL, p1 = 0.0294, p2 = 0.0245, sig.level = 0.05, power = 0.8,
                alternative = "two.sided")
```


**3. Verify R’s calculations for part 2) by hand using the method learned in class. Expect your
answer to be close in value to, but not exactly the same as, that provided by `R`, due to
rounding in hand calculations. (`Stata` uses a continuity correction that `R` doesn’t, so the
value calculated from `Stata` will also be different than the one you calculated by hand and in
R.)**


---

According to the information provided, $p_1 = 0.0294$, $p_2 = 0.0245$, $\alpha = 0.05$, $\beta = 0.2$,
$\Delta = 0.0245 * (1.2 - 1) = 0.0049$, $\bar{p} = \frac{p_1 + p_2}{2} = 0.02695$.

$n_1 = n_2 = \frac{(z_{\alpha/2}*\sqrt{2*\bar{p}*\bar{q}} + z_{\beta}*\sqrt{p_1*q_1 + p_2*q_2})^2}{\Delta^2} = \frac{(1.96*\sqrt{2*(0.02695)*(1-0.02695)} + 0.84*\sqrt{(0.0294)*(1-0.0294) + (0.0245)*(1-0.0245)})^2}{(0.0049)^2} = 17124.5 \approx 17125$.

---

To avoid differences caused by rounding issues, I used `qnorm()` function to conduct the calculation:

```{r}
(qnorm(0.025) * sqrt(2 * (0.02695) * (1 - 0.02695)) + qnorm(0.2) * sqrt((0.0294) *
(1 - 0.0294) + (0.0245) * (1 - 0.0245)))^2/(0.0049^2)
```

---
