---
title: "Problem Set 3"
author: "Zhengting (Johnathan) He"
date: "2022/3/5"
output: html_document
---


```{r "setup", include = FALSE}
require("knitr")
opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 3")
```


```{r}
setwd("D:/OneDrive - Johns Hopkins/Course/140.623.81 - Statistical Methods in Public Health III/Problem set/jhsphbiostat623-assignment/Problem Set 3")

require(tidyverse)
require(tidymodels)
require(poissonreg)
require(multcomp)
require(survival)
require(survminer)
```


**a. An alternative to calculating Kaplan-Meier estimates of the survival curve is to calculate *life-table estimates* when the time intervals are grouped or binned. Using the `lymphoma.csv`
data set, we could divide the total time of exposure into roughly ten bins and determine the
numbers of deaths and person-days experienced for each of the two groups in each bin. For
example, (0-7] is the bin more than 0 up to and including 7 days.**


**Assume the following bins: (0-7], (7-15], (15-30], (30-60], (60-90], (90-120], (120-150],
(150-180], (180-270], (270-360].**


**b. Download the `csv` data set `binlymph.csv` from CoursePlus. Verify that the calculations
of total time of exposure and person-days experienced appears to be correct by reviewing the
contents of this dataset. Also, using R create a plot of S(t) –vs.- mid_days for each group.**


```{r}
binData <- read_csv("./DATA/binlymph.csv")
binData %>% kable()
```


```{r}
qplot(x = mid_days, y = Survival,
      col = factor(stage, labels = c("Stage 3", "Stage 4")),
      data = binData) + geom_line() + labs(col = "Cancer Stage") + theme_minimal()
```


**c. Recall that D is the number of deaths, P_Days is the person-days accumulated in the bin and
mid_days is the midpoint of time bin. Rename variables for simplicity:**


```{r}
binData <- binData %>% 
    rename("t" = "mid_days", "N" = "P_Days")
```


**d. Fit the following four log-linear Poisson regression models to the grouped survival data**


|Model|$log \, EY_i$|
|--:|:--|
|A|$log \, N_i + \beta_0 + \beta_1 * stage$|
|B|$log \, N_i + \beta_0 + \beta_1 * stage + \beta_2 * (t-60)$|
|C|$log \, N_i + \beta_0 + \beta_1 * stage + \beta_2 * (t-60) + \beta_3 * (t-60)^+$|
|D|$log \, N_i + \beta_0 + \beta_1 * stage + \beta_2 * (t-60) + \beta_3 * (t-60)^+ + \beta_4*(t-60)*stage + \beta_5*(t-60)^+*stage$|


**e. Generate time terms, centered and spline:**


```{r}
binData <- binData %>% 
    mutate(t60 = t - 60) %>% 
    mutate(t60sp = ifelse(t > 60, t - 60, 0))
```


**f. Generate interaction terms: We don’t need to do this in R, since we can include the
interaction directly in our model.**


**g. Fit the models:**


```{r}
# modify tidy function
my_tidy <- function(x, exponentiate = FALSE, conf.level = 0.95, ...) {
    tidy <-
        dplyr::bind_cols(
            broom::tidy(x, conf.int = FALSE),
            # calculate the confidence intervals, and save them in a tibble
            stats::confint.default(x %>% extract_fit_engine()) %>%
                tibble::as_tibble() %>%
                rlang::set_names(c("conf.low", "conf.high"))
        )
    # exponentiating, if requested
    if (exponentiate == TRUE) {
        tidy <- tidy %>% dplyr::mutate_at(vars(estimate, conf.low, conf.high), exp)
    }
    tidy
}
```


```{r}
# poisson model specification
poisson_spec <- poisson_reg() %>% 
    set_engine("glm") %>% 
    set_mode("regression")
```


```{r}
# Model A: stage
modelA <- poisson_spec %>% 
    fit(D ~ stage + offset(log(N)), data = binData)
```


```{r}
# non exponentiate coefficients
modelA %>% my_tidy() %>% kable()
```


```{r}
# exponentiate coefficients - IRR
modelA %>% my_tidy(exponentiate = TRUE) %>% kable()
```


```{r}
# Model B: stage + t-60
modelB <- poisson_spec %>% 
    fit(D ~ stage + t60 + offset(log(N)), data = binData)
```


```{r}
# non exponentiate coefficients
modelB %>% my_tidy() %>% kable()
```


```{r}
# exponentiate coefficients - IRR
modelB %>% my_tidy(exponentiate = TRUE) %>% kable()
```


```{r}
# Model C: stage + t-60 + (t-60)^+
modelC <- poisson_spec %>% 
    fit(D ~ stage + t60 + t60sp + offset(log(N)), data = binData)
```


```{r}
# non exponentiate coefficients
modelC %>% my_tidy() %>% kable()
```


```{r}
# exponentiate coefficients - IRR
modelC %>% my_tidy(exponentiate = TRUE) %>% kable()
```


```{r}
# Model D: stage + t-60 + (t-60)^+ + stage*(t‑60) + stage*(t‑60)^+
modelD <- poisson_spec %>% 
    fit(D ~ stage + t60 + t60sp + stage:t60 + stage:t60sp + offset(log(N)), data = binData)
```


```{r}
# non exponentiate coefficients
modelD %>% my_tidy() %>% kable()
```


```{r}
# exponentiate coefficients - IRR
modelD %>% my_tidy(exponentiate = TRUE) %>% kable()
```


**h. Use the AIC = -2 log likelihood + 2(# of parameters) to identify the “best” prediction 
model from among A-D. Interpret the model results in a few sentences, as if for a journal 
article.**


```{r}
glance(modelA) %>% 
    bind_rows(glance(modelB)) %>% 
    bind_rows(glance(modelC)) %>% 
    bind_rows(glance(modelD)) %>% 
    kable()
```


```{r}
modelB_lincom <- glht(modelB %>% extract_fit_engine(), linfct = c("stage+t60 = 0"))
summary(modelB_lincom); exp(confint.default(modelB_lincom))
```


---

According to the `AIC` statistics, the "best" prediction model is modelB, with lowest `AIC = 52.9` in these 4 models. According to the output of modelB, the incidence rate ratio compared patients in stage 4 to stage 3 is 2.57 (95% CI: 1.09, 6.05), among patients under 60 years old. According to the output of linear combination of coefficients of modelB, the incidence rate ratio compared patients in stage 4 to stage 3 is 2.55 (95% CI: 1.08, 6.02), among patients age greater than 60 years old. Both incidence rate ratios are statistically significant under $\alpha = 0.05$.

---


**i. Now use the `csv` data set `lymphoma.csv`. Calculate *Kaplan-Meier (K-M) estimates* of 
the survival curve with 95% CI separately for each group. Plot the K-M curves against time.**


```{r}
lymphData <- read_csv("./DATA/lymphoma.csv")
head(lymphData) %>% kable()
```


```{r}
lymphData$SurvObj <- with(lymphData, Surv(days, died == 1))
km.stage <- survfit(SurvObj ~ stage, data = lymphData,
                    type = "kaplan-meier", conf.type = "log-log")
summary(km.stage)
```


```{r}
ggsurvplot(km.stage,
           data = lymphData,
           risk.table = TRUE,
           conf.int = TRUE,
           xlab = "time",
           ylab = "s(t)",
           title = "Kaplan-Meier survival estimates by cancer stage",
           legend.title = "Cancer Stage",
           legend.labs = c("Stage 3", "Stage 4"),
           palette = "jama")
```


**j. Compare the K-M curves versus the display of S(t) – vs- mid_days that you produced in 
steps a and b.**


---

The probability of survival overtime showed the same trend in both plots. The life-table method calculates the survival at the mid-point of each time intervals, and estimates the survival function using a line chart. The Kaplan-Meier method estimates the survival function at each event time through a "step" plot.

---


**k. Carry out a log-rank test and determine a p-value for the null hypothesis that the two 
population survival curves are the same for Stage 4 -vs- Stage 3 patients. What do you 
conclude?**


```{r}
survdiff(SurvObj ~ stage, data = lymphData)
```


---

According to the output of the log-rank test, the probabilities of survival over time are statistically significantly different between these 2 groups, under $\alpha = 0.05$.

---


**l. Fit a Cox proportional hazards model with an arbitrary baseline hazard and a group effect for 
stage.**


```{r}
model1 <- coxph(SurvObj ~ stage, data = lymphData, ties = "breslow")
model1 %>% tidy(exponentiate = TRUE, conf.int = TRUE) %>% kable()
```


**m. Compare the results of the log-rank test from part k. with the corresponding test for the Cox 
model in part l. Do they differ enough to change interpretation?**


---

According to the cox-regression output, the hazard ratio of death comparing patients in stage 4 to stage 3 is 2.61 (95% CI: 1.10, 6.18), and is statistically significant under $\alpha = 0.05$. The results suggest that both log-rank test and cox-regression showed significantly different probability of survival over time for patients in two stages, while cox-regression showed exactly which patient group has high probability of survival over time.

---


**n. Create an `R` script file that documents and archives the steps of your statistical analysis. This 
file will make your analysis "reproducible."**


**o. Summarize your findings as if for a public health journal.**


---

In this problem set, we compared the probability of survival over time comparing diffuse histiocytic lymphoma patients in stage 4 to stage 3.

We first treat the data as discrete time data, and divide the total time of exposure into ten binned time intervals, and used life-table method and log-linear Poisson regression to estimate the incidence rate ratio. According to the life-table estimates, patients in stage 3 have higher probability of survival than stage 4, over time. The crude incidence rate ratio compared patients in stage 4 to stage 3 is 2.98 (95% CI: 1.27, 6.97). The adjusted incidence rate ratio compared patients in stage 4 to stage 3 is 2.57 (95% CI: 1.09, 6.05), among patients under 60 years old, adjusting for age; and the adjusted incidence rate ratio compared patients in stage 4 to stage 3 is 2.55 (95% CI: 1.08, 6.02), among patients age greater than 60 years old, adjusting for age.

We then treat the data as continuous time data, and used Kaplan-Meier method and cox proportional hazard model to estimate the hazard ratio. According to the Kaplan-Meier method estimates, patients in stage 3 have higher probability of survival than stage 4, over time, and the difference is statistically significant under $\alpha = 0.05$. The crude hazard ratio compared patients in stage 4 to stage 3 is 2.61 (95% CI: 1.10, 6.18).

In summary, both methods suggest a statistically significantly lower probability of survival over time, compared diffuse histiocytic lymphoma patients in stage 4 to stage 3.

---

